\documentclass{llncs}

\usepackage{standalone}
\input{preamble.tex}
\input{macros.tex}

\title{%
	Automatically Translating Proof Systems
	for SMT Solvers to the λΠ-calculus%
	\thanks{Research reported in this document was supported
	by an Amazon Research Award,~2022.}}
\author{Ciarán Dunne\inst{1,3}, Guillaume Burel\inst{2,3}}
\institute{\scriptsize INRIA, ENS Paris-Saclay;
  \and ensIIE;
  \and SAMOVAR, Télécom SudParis, Institut Polytechnique de Paris, 91120 Palaiseau, France}

\begin{document}
\maketitle
% we're not just translating proofs, but also the proof
% systems embedded in Eunoia.

\begin{abstract}
	\noindent
	Eunoia is an emerging logical framework designed for
	specifying the proofs and proof systems of SMT solvers,
	namely \texttt{cvc5}.
	%
	We present a translation from Eunoia to the
	\emph{$λΠ$-calculus modulo rewriting} as implemented
	by the $\lambdapi$ proof assistant.
	%
	The translation is implemented by our tool \texttt{eo2lp},
	which we use for generating $\lambdapi$ encodings of
	%
	(a)~a large fragment of \texttt{cvc5}'s
	\textit{Cooperating Proof Calculus} (CPC), and
	%
	(b)~proofs produced by \texttt{cvc5} on \texttt{unsat}
	problems from various fragments of SMT-LIB.
\end{abstract}
%
The area of automated reasoning known as
\emph{satisfiability modulo theories} (SMT)
focuses on the development of tools for
checking the satisfiability of first-order
formulas over combinations of mathematical
theories~\cite{Barrett2021}.
%
% Whilst many  accomplish this task
% with similar strategies, the approach employed by
% a solver can vary, meaning that different solvers
% will excel in different theories.
%
Such tools (known as SMT \emph{solvers}) are
increasingly used as back-ends for various tasks:
hardware~\cite{DBLP:conf/fmcad/MattareiMBDHH18} and
software verification~\cite{barnett06boogie,riant23debugging,kosmatov16frama-c},
%
model checking \cite{DBLP:conf/cav/ChampionMST16},
cyber-security \cite{Backes2018},
and to improve automation in proof
assistants~\cite{DBLP:conf/cpp/ArmandFGKTW11,DBLP:journals/jar/BlanchetteBP13}.
%
Because these applications demand a high level
of confidence in the results produced by the solver,
many solvers are capable of generating
\emph{proof certificates}
that demonstrate the absurdity of specifications
deemed unsatisfiable by the solver.
%


Towards the aim of standardizing and benchmarking solvers,
the \emph{SMT library initiative} oversee the development of
the \emph{SMT-LIB standard} \cite{Barrett2015-standard},
which specifies the syntax and semantics of a
common language used for interacting with solvers.
%%
Proof generation is however not covered by the standard,
which has led to the development of various proof formats,
namely LFSC, Alethe, and recently Eunoia.
%
Eunoia is a proof output language and a logical framework:
it allows formalizing the constants, side-condition programs
and inference rules used by
% the proof generation facilities of
an SMT solver.
%
In particular, the proof system of $\texttt{cvc5}$ is
formalized as a Eunoia signature,
allowing its proofs to be checked by the C++ tool
Ethos~\cite{EthosUser_manualmdMain}.
%
%

% The LFSC format encodes inference rules in the
% Edinburgh Logical Framework (LF) extended with
% side-conditions,
% thereby reducing proof-checking to type-checking.
% % (i.e., executable programs that
% % may restrict applications of some inference rules),
% %
% In contrast, Alethe draws from a fixed set of
% inference rules that are designed to reflect
% the reasoning mechanisms of SMT solvers,
% %
% and proofs are checked by the standalone tool
% Carcara~\cite{andreottiCarcaraEfficientProof2023}.
% %

% Eunoia shares aspects with the speculative proposal
% for SMT-LIB 3~\cite{smt3-proposal}:
% it uses e.g.\ dependent types and binders.
% %
The $\lambda\Pi$-calculus modulo rewriting~\cite{CousineauDowek}
and its implementations in the tools Dedukti~\cite{assaf2023dedukti}
and LambdaPi~\cite{hondet:hal-02981561} are designed to
offer a trustworthy proof checker with an emphasis on
interoperability of proof systems.
%
This paper presents a translation from Eunoia to LambdaPi
and an OCaml tool $\texttt{eo2lp}$ that implements the
translation.
%
We used our tool to translate a large portion of
$\texttt{cvc5}$'s proof system to LambdaPi,
allowing us to further translate and typecheck
thousands of proof scripts produced by running $\texttt{cvc5}$
on $\texttt{unsat}$ problems from the SMT-LIB fragments for logic,
linear integer arithmetic, arrays, and sets.


%

% Coltellacci~\cite{coltellacciReconstructionSMTProofs2024a} developed a
% translation of Alethe proofs into LambdaPi, by adding a new back-end
% in the Carcara tool. Roughly, to each inference rule of the Alethe
% specification corresponds a lemma in LambdaPi, and lemmas are combined
% to reconstruct the proof.
\input{intro.tex}

\input{eoas.tex}

\input{lp.tex}

\input{res.tex}

\section{Conclusion and Future Work}\label{sec:conclu}

We have presented a translation from Eunoia to the
$\lambda\Pi$-calculus modulo rewriting, targeting the
logical framework that defines the proof system of the
SMT solver \cvc{5}.
%
Unlike approaches based on fixed proof formats such as
Alethe, our translation encodes the Eunoia framework
itself---including inference rule declarations,
side-condition programs, and elaboration
semantics---so that extensions to the proof system
require no changes to the translator.
%
Our tool \texttt{eo2lp} implements this translation and
successfully translates all 21~modules of CPC-MINI
(353~inference rules, 42~programs, 57~theory constants)
into $\lambdapi$ files that pass type-checking with
subject reduction verification.
%
On a benchmark of 1{,}227~proof scripts across 18~SMT-LIB
fragments, 91\% are translated and independently verified
by $\lambdapi$, with an average checking time of 0.47\,s.

Several directions for future work are possible.
%
First, extending CPC-MINI to the full CPC
would enable translation of arbitrary \cvc{5} proofs,
including those involving nonlinear arithmetic, strings,
and bitvectors---theories whose rules account for the
majority of current failures.
%
Second, integrating \texttt{eo2lp} into \cvc{5}'s
continuous integration pipeline would provide ongoing
independent verification of proof production.
%
Finally, the $\lambdapi$ encodings produced by
\texttt{eo2lp} can serve as a bridge to the Dedukti
ecosystem, enabling export of verified SMT proofs
to proof assistants such as Rocq and
Isabelle/HOL~\cite{DBLP:journals/corr/abs-1807-01873}.

\printbibliography

\end{document}


% \hrule
% \vspace{2mm}

% 	\hrule
% $$\begin{array}{r@{\ }l@{\quad}l}
% 		δ &
% 		\begin{array}[t]{c@{\ }l}
% 			{⦂} & \prn{\decl{x}{\prn{\vec ρ}}{\prn{{:}\ t}}{\prn{{∷}\ α}}}
% 			∣ \prn{\defn{x}{\paren{\vec ρ}}{\paren{{≔}\ t}}{\paren{{:}\ t'}}}    \\
% 			{∣} & \paren{\prog{x}
% 			{\paren{\vec ρ}}
% 			{\paren{{:}\ {\paren{t_1 \ldots t_n} → t'}}}
% 			{\paren{r_1 \ldots r_m}}
% 			}                                                                      \\
% 			{∣} & \paren{\irule{x}
% 				{\paren{\vec ρ}}
% 				{\paren{φ_1 \ldots φ_n}}
% 				{\paren{t_1 \ldots t_m}}
% 				{\paren{ψ}}
% 			}
% 		\end{array}
% 		  & \mcomment{(declarations)}
% 	\end{array}
% $$
