\documentclass[class=llncs, crop=false]{standalone}

\input{preamble.tex}
\input{macros.tex}

\begin{document}
% ---------------------------------------------------------
\begin{boxfigure}[t!]{fig:eo-syntax}
	{Syntax for Eunoia: terms, attributes, and commands.}
	%
	$$ \eoTermSyntax $$
	\hrule
	$$ \eoAttrSyntax $$
	\hrule
	$$ \eoCommandSyntax $$
	\hrule
	$$ \eoPrfCommandSyntax $$
\end{boxfigure}
% ---------------------------------------------------------
\Autoref{fig:eo-syntax} defines syntax for the
fragment Eunoia considered in this document, where the
metavariable $s$ ranges over a set of \emph{symbols}.
%
First, we define grammars for the core term syntax.
%
Each term is either a symbol $s$, an (integer or string)
\emph{literal} $ℓ$, an \emph{application} $\eapp{s}{\vec t}$,
or a \emph{binder application} $\eapp{s}{\epar{\vec ρ}\ t}$.
%
Each \emph{parameter} $ρ$ consists of a symbol $s$, a term
% forward ref to where we discuss typing ??
$t$ (the type of the parameter), and possibly a
\emph{parameter attribute} $ν$.
%

Second, we define the \emph{commands} of Eunoia, which are
divided into \emph{standard commands} and \emph{proof commands}.
%
Hereinafter, a Eunoia \emph{signature} is a list $Δ$
of standard commands, and a \emph{proof script} is a
list $Υ = \prn{\vec δ ⨾ \vec π}$
where $\vec δ$ is a list of standard commands called
the \emph{preamble} and $\vec π$ is a list of
proof commands called the \emph{body} of the script.
%
In practice, a Eunoia-friendly SMT solver should provide a
signature $Δ$ that encodes the inference rules that may be
used in the proof scripts produced by the solver.
%
Then, the correctness of a proof script $Υ$ (wrt. $Δ$)
may be checked using the Ethos checker.
%

\subsection{Commands and their Declarations}\label{sec:eo-judge}
%
We define an abstract interface for reasoninga about
signatures and proof scripts.
%
This interface is useful for
characterizing the \emph{elaboration}
and \emph{translation} operators
found in \autoref{sec:eo-elab} and
\autoref{sec:results} respectively.

\subsubsection{Constant Declaration.}
%
Let $δ$ be a \emph{(parameterized) constant declaration}
with symbol $s$, parameters $\vec ρ$, and term $t$.
%
We may write $\prn{⊢ s(\vec ρ) : t}$ to express
that the command $δ$ declares $t$ as the \emph{type} of $s$
with respect to parameters $\vec ρ$.
%
Furthermore, if a constant attribute $α$ is given by $δ$,
we may write $\prn{δ ⊢ s(\vec ρ) ∷ α}$.
%

\subsubsection{Macro Definition.}
%
Let $δ$ be a \emph{macro definition} with
symbol $s$, parameters $\vec ρ$, and term $t$.
%
Then $δ$ declares $t$ as the \emph{definiens} of $s$
wrt. $\vec ρ$; written $\prn{δ ⊢ s(\vec ρ) ≔ t}$.
%
If the attribute $\mtt{:type}\ t'$ is given by $δ$,
then $\prn{δ ⊢ s(\vec ρ) : t'}$ also holds.

\subsubsection{Program Declaration.}
%
Let $δ$ be a \emph{program declaration} with
symbol $s$, parameters $\vec ρ$,
$\mtt{:signature}\,\prn{\plur t n}\,t'$,
and cases $\plur c m$.
%
Then, $δ$ declares the type and definiens of $s$
as follows,
where $\plur t n$ are the \emph{domain} types of $s$
and $t'$
is \emph{range}:
% Let $\vec t = \plur t n$ and $\vec r = \plur r m$,
% and $δ ∈ Δ$.
% $(\plurcomma t n)$, \emph{range} $t'$ and
% \emph{cases} $\vec r$ with respect to $\vec ρ$.
%
% More formally, we may write:
$$
	δ ⊢ s(\vec ρ) : \earr{\plur t n\, t'}
	\quad\text{and}\quad
	δ ⊢ s(\vec ρ) ≔ \mbf{cases}[c_1, \ldots, c_m]
$$
% for the type of $s$,
% and
%   $$
% for the cases.
%
\subsubsection{Inference Rule Declaration.}
%
Let $δ$ be a \emph{rule declaration} with symbol $s$,
parameters $\vec ρ$, and \emph{conclusion} $φ$.
%
Also, suppose $δ$ provides
$\mtt{:premises}\ \epar{\plur ψ n}$,
$\mtt{:args}\ \epar{\plur t m}$ with types $\plur τ m$,
and $\mtt{:requires}\
	\epar{\epar{x_1\ y_1}\ldots\epar{x_o\ y_o}}$.
%
First, let $δ$ declare the type and definiens
of an \emph{auxiliary symbol} for $s$ thus:
$$
	\begin{array}[t]{l@{\ ⊢\ }l}
		δ & s^⭑(\vec ρ) : \earr{\plur τ m\ \bool}
		\\[1mm]
		δ & s^⭑(\vec ρ) ≔ \mbf{cases}
		[\;
			\epar{{\eapp{s^⭑}{\plur t m}}\ φ}
			∣
			\epar{x_1\ y_1}\ldots\epar{x_o\ y_o}
			\;]
	\end{array}
$$
%
Then, the type of $s$ is given by $δ$ with an extended list of parameters%
\footnote{The symbols chosen for $\plur α m$ must be \emph{fresh} with respect to $δ$.
	That is, each $α_i$ is distinct from any symbol occurring in any of the
	terms or parameters supplied by $δ$.}
thus:
$$δ ⊢ s(\vec ρ, \epar{α_1\ τ_1} \ldots \epar{α_n\ τ_m}) :
	\earr{
		\epar{\mtt{Proof}\ ψ_1} \ldots
		\epar{\mtt{Proof}\ ψ_n}\
		\epar{\mtt{Proof}\ φ^⭑}
	}$$
where $φ^⭑ ≔ \eapp{s^⭑}{\plur α n}$
ensures that a term of type $\eapp{\mtt{Proof}}{φ}$
may only be obtained iff $\prn{α_i = t_i}$
for $1 ≤ i ≤ m$ and $\prn{x_j = y_j}$ for $1 ≤ j ≤ o$.
% and $$.

\subsubsection{Signature Inclusion.}
%
Hereinafter, let $Θ$ be a \emph{(global) environment}
mapping from filepaths to Eunoia signatures.
%
Let $δ$ be an inclusion with valid filepath $μ$.
Then, any judgement $J$ made by a command $δ'$ in $Θ_μ$
is also made by $δ$.
That is:
$$
	\text{$∀\,δ' ∈ Θ_μ$,}\quadδ' ⊢ J \quad⟹\quad \epar{\mtt{include}\ μ} ⊢ J
$$

\subsubsection{Proof Scripts.}

Two basic forms of \emph{proof commands} are given,
which are called \emph{assumption} and \emph{step}
respectively.
%
Given $π = \epar{\mtt{assume}\ s\ φ}$ or
$π = \epar{\mtt{step}\ s\ φ\ \ldots}$, we call
$s$ the \emph{name} of $π$ and $φ$ the \emph{conclusion}.
%
In either case, we may write:
$$π ⊢ s : \eapp{\mtt{Proof}}{φ}$$
%
Furthermore, let $π$ be a step with
$\mtt{:rule}\ s'$,
$\mtt{:premises}\ \epar{\vec p}$, and
$\mtt{:args}\ \epar{\vec t}$.
%
Then, $π$ judges the definiens of $s$
as the application of $s'$ to the premises
$\plur p n$ and arguments $\plur t m$, i.e.;
$$π ⊢ s ≔ \eapp{s'}{\plur t m\ \plur p n}$$
%



% iff $π ∈ Π$.

%
% We define We say that a proof scripts is \emph{valid} with respect
% to $Δ$ iff a list of
% commands of the form $\vec δ, \vec π$ proof commands.



% ----------------------------------------------------------
\subsection{Elaboration of Terms}\label{sec:eo-elab}
%
In Eunoia, the `de-sugared' meaning of an application
$\eapp{s}{\vec t}$ depends on the constant attribute
assigned to $s$ within some signature $Δ$.
%
For example, consider a constant declaration $δ$ with:
$$
	δ ⊢ \eor : \earr{\bool\ \bool\ \bool}
	\quad\text{and}\quad
	δ ⊢ \eor ∷ {\rcn \false}
$$
% and $δ ⊢ $.
% $$\dca{\eor}{\earr{\bool\ \bool\ \bool}}$$
%
From the type of $\eor$, we may expect its only valid
uses to be of the form $\eapp{\eor}{t_1\ t_2}$.
%
However, the assignment of the constant attribute means
that the $\mtt{or}$ symbol is treated as
\emph{right-associative} with \emph{nil-terminator} $\false$.
%
Thus, $n$-ary applications of $\eor$ are
\emph{elaborated} to a normal form, e.g.;
%
$$\eapp{\eor}{\ev x\ \ev y\ \ev z}\ ⇢\
	\eapp{\eor}{\ev x\ \eapp{\eor}{\ev y\ \eapp{\eor}{\ev z\ \false}}}$$
%
Furthermore, the elaboration of such applications may
also depend on the attributes of `locally bound' symbols.
For example, consider $\vec ρ$ containing
parameters $\epar{\ev x\ \bool}$,
$\epar{\ev y\ \bool\ \mtt{:list}}$,
and $\epar{\ev z\ \bool}$.
%
Then, observe:
$$
	\eapp{\eor}{\ev x\ \ev y\ \ev z}
	\ {⇢}
	\ \econs\eor{\ev x}{
		\econcat{\eor}{\ev y}{\eapp{\eor}{\ev z\ \false}}
	}
$$
%
Here, $\eo{list\_concat}$ is a built-in ternary operator:
$\eapp{\eo{list\_concat}}{f\ \ell\ \ell'}$ concatenates
two $f$-lists $\ell$ and $\ell'$.
%
The $\mtt{:list}$ attribute alters the elaboration strategy
under the assumption that $\ev y$ will (eventually) be
substituted for some $\eor$-list.
%
It may be helpful for the reader
to consider the result of substituting
$\ev y ↦ \econs\eor{\ev w}{\false}$
thus:
$$
	\begin{array}[t]{c}
		\econs\eor{\ev x}{
			\econcat{\eor}{\econs\eor{\ev w}{\false}}
			{\eapp{\eor}{\ev z\ \false}}}
		\\[1mm] {⇣} \\[1mm]
		\econs\eor{\ev x}{
			\econs\eor{\ev w}{
				\econs\eor{\ev z}{\false}}
		}
	\end{array}
$$
%
With the aim of supporting the elaboration strategies
corresponding to the constant attributes given in
\autoref{fig:eo-syntax},
we define an \emph{elaboration} operator below.
%
\begin{definition}
	For any symbol $s$ and list of parameters $\vec ρ$,
	let $\glue{\vec ρ,f}$ be the binary operator
	such that for any terms $t_1$, $t_2$, the following holds:
	%
	$$
		\glue{\vec ρ,f}[t_1,t_2] =
		\begin{cases}
			\econcat{f}{t_1}{t_2} & \text{if $\vec ρ ⊢ t_1 ∷ \mtt{:list}$,}
			\\
			\econs{f}{t_1}{t_2}   & \text{otherwise.}
		\end{cases}
	$$
	%
	Then for any signature $Δ$, let $\elab{Δ,\vec ρ}$
	be the least (unary) operator such that:
	\begin{enumerate}
		\item For any symbol $s$, $\elab{Δ,\vec ρ}[s] = s$.
		      %
		\item Let $s$ be a symbol with $Δ ⊢ s ∷ \bd t_\cons$
		      (for some term $t_\cons$), and
		      let $\vec v = \epar{x_1\ t_1}\ldots\epar{x_n\ t_n}$.
		      Then for any term $t$, we have:
		      $$ \elab{Δ,\vec ρ}[\eapp{s}{\epar{\vec v}\ t}]
			      =
			      \eapp{s}{t_{\text{vars}}\ t_{\text{body}}}
		      $$
		      where
		      $t_{\text{vars}} ≔ \elab{Δ, \vec ρ}
			      [
				      \eapp{t_\cons}{
					      \eovar{\mtt{"}x_1\mtt{"}}{t_1}
					      \ldots
					      \eovar{\mtt{"}x_n\mtt{"}}{τ_n}
				      }
			      ]
		      $
		      encodes $\vec v$ as a list of $\eo{var}$ terms,
		      and $t_{\text{body}} ≔ \elab{Δ, \vec ρ}[t']$
		      where $t'$ is the result of replacing each
		      symbol $x_i$ in $t$ with its $\eo{var}$ encoding.
		      %
		\item For any symbol $s$ and list of terms $\vec t$:
		      $$
			      \begin{array}[t]{l}
				      \elab{Δ,\vec ρ}[\eapp{s}{\vec t}]\ {=}\
				      \\[2mm]
				      \quad
				      \begin{cases}
					      \foldr(G, t_\nil, \vec t')
					       & \tsm{if $Δ ⊢ s ∷ \rcn{t_\nil}$, }
					      \\
					      \foldl(G', t_\nil, \vec t')
					       & \tsm{if $Δ ⊢ s ∷ \lcn{t_\nil}$,}
					      \\
					      \foldr(G, t'_n, t'_1\ldots t'_{n-1})
					       &
					      \tsm{if $Δ ⊢ s ∷ \rc$,}
					      \\
					      \foldl(G', t'_1, t'_2\ldots t'_{n})
					       &
					      \tsm{if $Δ ⊢ s ∷ \lc$,}
					      \\
					      \elab{Δ,\vec ρ}[\eapp{{t_\cons}}{\mbf{chain}(s, \vec t')}]
					       &
					      \tsm{if $Δ ⊢ s ∷ \ch {t_\cons}$,}
					      \\
					      \elab{Δ,\vec ρ}[\eapp{{t_\cons}}{\mbf{pairs}(s, \vec t')}]
					       &
					      \tsm{if $Δ ⊢ s ∷ \pw {t_\cons}$,}
					      \\
					      \eapp{s}{{t_1'}\ldots{t_n'}}
					       &
					      \tsm{otherwise.}
				      \end{cases}
				      \\[3mm]
			      \end{array}
		      $$
		      where\footnote{%
			      $\foldl(G, a, l)$ (resp.\ $\foldr(G, a, l)$) is
			      the standard left fold (resp.\ right fold) using
			      combining function $G$ and accumulator $a$ on list $l$.
		      } $G ≔ \glue{\vec ρ, f}[x,y]$,
		      $G'(x,y) ≔ G(y,x)$,
		      and $t_i' ≔ \elab{Δ, \vec ρ}[t_i]$, and:
		      $$
			      \begin{array}[t]{r@{\ ≔\ }l}
				      \mbf{chain}(f,\, t_1\ldots t_n)
				       & \eapp{f}{t_1\ t_2}\,
				      \eapp{f}{t_2\ t_3}
				      \,\ldots\,
				      \eapp{f}{t_{n-1}\ t_n}
				      \\[1mm]
				      \mbf{pairs}(f,\, t_1\ldots t_n)
				       & \eapp{f}{t_1\ t_2}
				      \,\ldots\,
				      \eapp{f}{t_1\ t_n}\,
				      \eapp{f}{t_2\ t_3}
				      \,\ldots\,
				      \eapp{f}{t_{n-1}\ t_n}
			      \end{array}
		      $$
	\end{enumerate}

	%


\end{definition}

%













% ----------------------------------------------------------


\end{document}


% \subsection{Terms, Types and Attributes}
% %
% The \emph{built-in constants} of Eunoia are given by a fixed
% subset of symbols and denoted by the following expressions:
% $\mtt{Type}$ (the kind of all types),
% $\mtt{->}$ (arrow-type constructor),
% $\mtt{\_}$ (binary function application),
% $\mtt{Bool}$ (the type of Boolean values),
% $\mtt{true}$ and $\mtt{false}$ (terms of type \mtt{Bool}).
