\documentclass[class=llncs, crop=false]{standalone}
\input{preamble.tex}
\input{macros.tex}

\begin{document}
\section{Introduction}\label{sec:intro}
%
The area of automated reasoning known as
\emph{satisfiability modulo theories} (SMT)
focuses on the development of tools for
checking the satisfiability of first-order
formulas over combinations of mathematical
theories~\cite{Barrett2021}.
%
Such tools (known as SMT \emph{solvers}) are
increasingly used as back-ends for various tasks:
hardware~\cite{DBLP:conf/fmcad/MattareiMBDHH18} and
software verification~\cite{barnett06boogie,riant23debugging,kosmatov16frama-c},
%
model checking \cite{DBLP:conf/cav/ChampionMST16},
cyber-security \cite{Backes2018,rungta22billion},
and to improve automation in proof
assistants~\cite{DBLP:conf/cpp/ArmandFGKTW11,DBLP:journals/jar/BlanchetteBP13}.
%
Because these applications demand a high level
of confidence in the results produced by the solver,
many solvers are capable of generating
\emph{proof certificates}
that witness the unsatisfiability of formulas
reported as such by the solver.
%

% Towards the aim of standardizing and benchmarking solvers,
% the \emph{SMT library initiative} oversees the development of
The \emph{SMT-LIB standard}~\cite{Barrett2015-standard}
specifies the syntax and semantics of a
common language used for interacting with solvers.
%%
Proof generation is however not covered by the standard,
leading to the development of various proof formats,
namely LFSC~\cite{DBLP:journals/fmsd/StumpORHT13},
Alethe~\cite{schurrAletheGenericSMT2021},
and recently Eunoia~\cite{EthosUser_manualmdMain}.
%
Eunoia is both a proof output language and a logical framework:
it allows specifying the constants, side-condition programs,
and inference rules of an SMT solver (e.g., \cvc{5}).
The \emph{Cooperating Proof Calculus} (CPC) is the
Eunoia signature encoding \cvc{5}'s proof system,
and Eunoia proof scripts may be checked by the C++ tool
Ethos~\cite{EthosUser_manualmdMain}.
%
%

The $\lambda\Pi$-calculus modulo rewriting~\cite{CousineauDowek}
and its implementations in the tools Dedukti~\cite{assaf2023dedukti}
and LambdaPi~\cite{hondet:hal-02981561} are designed to
offer a trustworthy proof checker with an emphasis on
interoperability of proof systems.
%
This paper presents a translation from Eunoia to LambdaPi
and an OCaml tool $\texttt{eo2lp}$ that implements the
translation.
%
We used our tool to translate a large portion of
\cvc{5}'s proof system to LambdaPi (\autoref{fig:pipeline}),
allowing us to further translate and typecheck
over 1{,}000~proof scripts produced by running \cvc{5}
on $\texttt{unsat}$ problems from the SMT-LIB fragments for logic,
linear integer arithmetic, arrays, and sets.
% ---------------------------------------------------------
\begin{boxfigure}[t!]{fig:pipeline}
	{{Overview of the translation pipeline.
				\cvc{5} produces Eunoia proof scripts using rules
				from the CPC signature, which may be checked by Ethos.
				\texttt{eo2lp} translates both the signature and the proofs
				to LambdaPi, which independently verifies them.}}
	%
	\centering
	\alttext{Pipeline diagram with three vertical groups.
		Left: SMT-LIB fragments (QF\_UF, QF\_LIA, UFLIA, ALIA, QF\_NIA, UF, \ldots).
		Middle: Eunoia artefacts---a CPC signature (theories, rules, programs)
		and a Eunoia proof script (.prf.eo).
		Right: LambdaPi artefacts---a translated CPC signature plus Prelude,
		and a LambdaPi proof file (.prf.lp).
		Horizontal arrows: cvc5 maps SMT-LIB fragments to Eunoia proof scripts;
		eo2lp maps both the Eunoia signature and proof script to LambdaPi.
		Downward arrows: ethos checks the Eunoia bundle, lambdapi check verifies
		the LambdaPi bundle, each producing a tick-or-cross result.}{%
		\input{fig-pipeline.tex}}
\end{boxfigure}
% ---------------------------------------------------------

\subsubsection{Contribution.}
Our work has a dual role.
Practically, \texttt{eo2lp} (\autoref{sec:results})
provides an independent type-theoretic proof checker
for \cvc{5}, translating all 22~modules of CPC-MINI
(359~rules, 96~programs, 57~constants) and verifying
97\% of 1{,}036~proof scripts across 14~SMT-LIB fragments.
%
Conceptually, the translation gives a
\emph{typed interpretation} of Eunoia in the
$λΠ$-calculus, where Eunoia sorts become elements of a
Tarski universe~$\Set$ with carrier~$\msf{τ}$,
inference rules become axioms, and side-condition
programs become rewrite rules whose type preservation
is checked by LambdaPi.
%
This parallels SMT-LIB's model-theoretic semantics
where sorts are interpreted as sets, and is a step
towards a formal semantics for the still-evolving
Eunoia language.

The paper is organised as follows.
\Autoref{sec:eunoia} formalizes the fragment of
Eunoia used by \cvc{5}---its term syntax, command
structure, and elaboration semantics.
%
\Autoref{sec:lambdapi} defines the translation
of Eunoia into the $λΠ$-calculus modulo rewriting,
covering inference rules, side-condition programs,
and elaboration uniformly so that \cvc{5}'s entire
proof system is encoded automatically rather than
rule-by-rule.
%
\Autoref{sec:results} describes \texttt{eo2lp}
and reports the benchmark results.

\subsubsection{Related Work.}
Coltellacci et
al.~\cite{coltellacciReconstructionSMTProofs2024a,coltellacci25frocos}
reconstruct Alethe~\cite{schurrAletheGenericSMT2021}
proofs from
\cvc{5}~\cite{Barbosa2022}
in LambdaPi via hand-written soundness lemmas for each
Alethe rule, covering UF and linear integer arithmetic.
%
Similarly, \textsc{lean-smt}~\cite{mohamed25leansmt}
covers ${\sim}200$ of the 662~CPC rules
(${\sim}30\%$) by interactively proving each rule
sound in Lean~4, then automating the translation
of cvc5 proofs that use them.
%
Our work instead targets Eunoia as a framework:
all 359~CPC-MINI rules are translated automatically as
\emph{axioms} in LambdaPi, without per-rule soundness
proofs, but whose well-typedness and subject reduction
are verified.

The idea of reducing SMT proof checking to type
checking originates with
LFSC~\cite{DBLP:journals/fmsd/StumpORHT13};
Eunoia inherits this approach and extends it with
the higher-order features similar to those of
the speculative SMT-LIB~3
proposal~\cite{barrettSMTLIB3Bringing,smt3-proposal,barbosa17hosmt}.
%
Brown~\cite{Brown2021} argues that this expressive
type system calls for proof representations in
type-theoretic frameworks; our encoding realizes
this idea using the $λΠ$-calculus.
%
SMT proof reconstruction is also used within
proof assistants:
Isabelle/HOL replays Alethe
proofs~\cite{DBLP:conf/cade/SchurrFD21,lachnitt25isabelle},
SMTCoq~\cite{ekici17smtcoq} integrates solvers
into Rocq, and
IsaRare~\cite{lachnitt24isarare} verifies
CPC rewrite rules in Isabelle/HOL.

The Dedukti ecosystem serves as an
interoperability hub for proof assistants
(Rocq~\cite{boespflug12coqine},
Isabelle~\cite{isabelle_dedukti},
HOL~Light~\cite{assaf15translating},
Matita~\cite{krajono},
Lean~\cite{vaishnav25lean4less})
and automated provers
(iProverModulo~\cite{burel20first},
ArchSAT~\cite{bury:tel-02612985},
Vampire~\cite{komel2025case},
TPTP~\cite{DBLP:conf/flairs/SutcliffeBB25});
our work extends this ecosystem to SMT via
the Eunoia framework.

% ---------------------------------------------------------
% Placed here so it floats to the top of page 2
\begin{boxfigure}[t!]{fig:eo-syntax}
	{Syntax for Eunoia: terms, attributes, and built-in operators.}
	\alttext{Three stacked grammar tables for Eunoia.
		Top: term syntax---terms are symbols, literals, applications, or
		binder applications; parameters carry an optional attribute (:implicit, :list).
		Middle: constant attributes---right/left-associative (with optional nil terminator),
		chainable, pairwise, binder, arg-list.
		Bottom: built-in symbols grouped by family---core (eo::eq, eo::requires, eo::quote,
		eo::var), boolean (eo::and, eo::or, eo::not, eo::xor),
		arithmetic (eo::add, eo::mul, eo::neg, eo::zdiv), and
		list (eo::nil, eo::cons, eo::list\_concat, eo::list\_len).}{%
		\vspace{4pt}\centerline{$\eoTermSyntax$}
		\boxfigurehline
		\centerline{$\eoAttrSyntax$}
		\boxfigurehline
		\centerline{$\eoBuiltinFig$}\vspace{4pt}}
\end{boxfigure}
% ---------------------------------------------------------

\end{document}
