\documentclass[class=llncs, crop=false]{standalone}

\input{preamble.tex}
\input{macros.tex}

\begin{document}
\section{Implementation and Results}\label{sec:results}

Our tool
\texttt{eo2lp}\footnote{\url{https://github.com/deducteam/eo2lp}}
is an OCaml program
(${\sim}5{,}500$~lines) that implements
\autoref{sec:lambdapi}, with a Menhir-generated
parser~\cite{menhir} and a back-end built against
the LambdaPi OCaml API.
%
A hand-written \texttt{Prelude.lp} module
(${\sim}335$~lines) bootstraps the type embedding:
it declares $\Set$, $\msf{τ}$, the function space~$⤳$,
the higher-order application symbol $\lpo{app}$,
Boolean logic, and arithmetic builtins as rewrite
rules over LambdaPi's integers/rationals
(e.g.\ $\eo{add}\,(\ev{of\_Z}\,i)\,(\ev{of\_Z}\,j) ↪
	\ev{of\_Z}\,(i + j)$).
The prelude also defines a \emph{heterogeneously-typed}
list type
$\eo{hlist} : \Set$ whose entries carry their own
sort, used to encode the
$\eapp{\eo{var}}{\estr{x_i}\ T_i}$ variables produced
by $\mbf{elab}$ on binder applications
(\autoref{sec:eo-elab}, item~2).
Closed expressions reduce concretely, so CPC's
\texttt{\$run\_evaluate} program---which recursively
delegates to the $\eo{add}, \eo{mul}, \eo{and},
	\eo{not},\ldots$ builtins---normalises ground terms
like $\eapp{\ev{+}}{2\ 3}$ to $5$.

\subsection{CPC-MINI}
The Cooperating Proof Calculus (CPC) is the
Eunoia signature that formalizes \cvc{5}'s proof
system.
%
The full CPC comprises 35~core modules
(${\sim}11{,}000$~lines of Eunoia) plus
16~expert modules for additional theories
(separation logic, bags, finite fields,
floating points, transcendentals).
%
To validate our approach on a manageable portion,
we define
\emph{CPC-MINI},\footnote{\url{https://github.com/ciaran-matthew-dunne/cpc-mini}}
a fork of the upstream CPC included as a git
submodule of \texttt{eo2lp}.
%
It is a modified subset of
22~core modules (${\sim}5{,}000$~lines)
covering logic, integer arithmetic, arrays,
sets, quantifiers, and uninterpreted functions
--- 6~theory modules (57~constants),
8~rule modules (353~rules, 50~programs),
and 7~program modules (38~programs, 36~macros),
plus a root module \texttt{Cpc}.

CPC-MINI applies three classes of change to the
upstream CPC.
\emph{(i)~Theory pruning:} the expert modules and
unsupported core theories (bitvectors, strings,
datatypes, reals) are removed, since their primitives
are not yet in our prelude.
\emph{(ii)~Type-parameter tightening:} a few CPC
declarations leave their numeric parameter type
unconstrained so that the same rule applies at both
\texttt{Int} and \texttt{Real}; we instantiate these
to \texttt{Int} where required, which suffices for
our integer-arithmetic benchmarks.
\emph{(iii)~Bug fixes:} several upstream typos and
mismatched type parameters are patched; we intend to
contribute these upstream. For instance, the
upstream \texttt{\$mk\_cong\_rhs} program in
\texttt{Uf.eo} typed its arguments with the codomain
rather than the domain of its function
parameter---a mismatch Ethos missed because it only
typechecks program clauses at application time.
%
All 22~modules pass
\texttt{lambdapi check} with full subject-reduction
verification; the entire CPC-MINI signature is
translated in ${\sim}100$\,ms and type-checked in
${\sim}160$\,ms---this is one-off work, independent
of the proofs to be verified later.

\subsection{Proof Benchmarks}
%
To evaluate the translation on proofs,
we sampled up to 100~unsatisfiable problems from
SMT-LIB for each of the 14~fragments covered by
CPC-MINI and obtained Eunoia proof scripts from
\cvc{5}, yielding 1{,}036~proofs.
%
Each proof is translated by \texttt{eo2lp} and
the result is checked by \texttt{lambdapi check}
with subject-reduction verification,
under a 5-second timeout.
%
\Autoref{tab:results} summarizes the results:
1{,}002 of 1{,}036~proofs (97\%) are
successfully translated and verified,
with an average checking time of 0.47\,s.

\input{results-table.tex}

The seven equality and array fragments achieve a
near-perfect pass rate (582 of 583), with the single
failure being a timeout. The seven linear-arithmetic
fragments pass 420 of 453 proofs (93\%):
QF\_UFIDL, UFLIA, and LIA reach 98--100\%; the
remaining failures are 18 timeouts in QF\_LIA and
QF\_IDL where large proofs exceed the 5-second
limit, and 15 type-checking errors in QF\_UFLIA
caused by CPC rules whose type signatures mix
\texttt{Real} and \texttt{Int}---most prominently
\texttt{arith-int-geq-tighten}, which declares a
\texttt{Real} parameter but is applied here in an
integer-only context.
Such mixed-arithmetic rules are excluded by
CPC-MINI's type-parameter tightening (change~(ii)
above); the failure is therefore a \emph{type
	coverage} gap at the \texttt{Real}/\texttt{Int}
boundary, not a translation bug. Crucially,
\emph{zero} failures are caused by a semantic
disagreement with Ethos: every proof that Ethos
accepts and we translate is accepted by LambdaPi up
to coverage and time.

To probe scaling, we ran the translator on a larger
sample of 439 \texttt{QF\_UF} proofs (up to
${\sim}5{,}000$ lines each) under a 30-second
timeout: all 439 check successfully. The encoding's
overhead over pure type-checking is one $\msf{τ}$
indirection per type plus rewriting for program
evaluation; pure type-checking is linear, but the
rewrite-based program evaluation needed for
side-conditions and arithmetic can blow up on
unusually long chains. Outliers in the main
benchmark are dominated by such chains, not by
translation bugs.

\end{document}
